D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-07 ootavacuous.tex
FYI, possible progress on language-level OOTA problem
Some more instructive examples of semantic dependencies.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

We have seen several examples showing that semantic dependencies may vary
according to the execution.
This raises the question: What exactly is an execution?

@@

The implementation-defined aspects can affect whether or not an
abstract execution contains a semantic dependency.
As an example consider the following, where the type of \co{c} is \co{char}:
\begin{quote}
\begin{verbatim}
y = x * (c >= 0);
\end{verbatim}
\end{quote}
Here \co{y} is semantically dependent on \co{c} in executions
for which \co{char} is a signed type,
but not those for which it is unsigned.

@@

\begin{enumerate}
\item	Some hardware might choose to forego ECC, for example, in order to
	reduce cost for low-end systems or to improve energy efficiency
	for battery-powered systems.
\item	ECC error correction might still be done in hardware, avoiding
	the need for precise software-visible exceptions.
\item	Some systems might prefer to immediately shut down, forbidding
	software from continuing after an uncorrectable ECC error,
	perhaps due to safety considerations.
\item	Someone might come up with a clever way of correcting ECC errors
	in firmware, avoiding the need for precise software-visible
	exceptions.
\item	As late as early 2024, some GPGPUs have been observed reordering
	earlier loads against later stores.
	Other hardware architects might therefore feel the need to keep
	this option open for their own systems.
\end{enumerate}

>>>>

\subsubsection{Semantic Dependencies Not Affected by \co{if} Statements}
\label{sec:Semantic Dependencies Not Affected by if Statements}

Compare this example to the previous one:
\begin{quote}
\begin{verbatim}
if (x > 0)
    y = 42;
else
    y = 42;
\end{verbatim}
\end{quote}
Because the stores executed on each arm of the \co{if} statement write
identical values to identical addresses, one could equally well regard
the two statements as performing two different stores or as performing
for all intents and purposes a single store, independent of x.
Reasonable C++ implementations might disagree on this matter and
therefore on whether or not the example has a semantic dependency.
It is the implementations's choice.

\subsubsection{Semantic Dependencies and Matching Up Stores}
\label{sec:Semantic Dependencies and Matching Up Stores}

Suppose we take the view that the previous example involves only one
store.
This opens up the door to greater complexity:
\begin{quote}
\begin{verbatim}
if (x > 0) {
    L1: y = 42;
} else {
    y = 53;
    y = 42;
}
\end{verbatim}
\end{quote}
Consider an execution in which \co{x} is greater than zero, so the
statement labeled \co{L1} runs.
Is it semantically dependent on \co{x}?
The answer isn't immediately clear.
If the other arm of the \co{if} is taken then the same store of 42 to
\co{y} occurs, but 53 is written before it.
Which of these two stores should be compared with the store in \co{L1}?

One way to cut the Gordian knot is to match up the stores by the order
they occur:
Since \co{L1} is the first store to \co{y} in its arm of the \co{if}
statement, it should be matched up with the first store to \co{y} in
the other arm.
Those two stores write different values so there is a semantic
dependency.

On the other hand, a compiler may decide to drop the \co{y =
53} store entirely, leaving it out of the machine code,
on the grounds that it's always possible for the two adjacent stores
to \co{y} to execute in such quick succession that no other thread
manages to read the value 53 before it gets overwritten with 42.
If the compiler does this then the first store to \co{y} in that
arm of the \co{if} statement \emph{would} agree with the store in
\co{L1}, and so there would not be a semantic dependency.
Once again, the decision is up to the implementation.

We have seen several examples showing that semantic dependencies may vary
according to the execution and even the implementation.
This raises some questions, of which the first is:
What exactly is an execution?

@@

The implementation-defined aspects can affect whether or not an
abstract execution contains a semantic dependency.
As an example consider the following, where the type of \co{c} is \co{char}:
\begin{quote}
\begin{verbatim}
y = (c >= 0);
\end{verbatim}
\end{quote}
Here \co{y} is semantically dependent on \co{c} in executions
for which \co{char} is a signed type,
but not those for which it is unsigned.

@@

\begin{enumerate}
\item	Some hardware might choose to forego ECC, for example, in order to
	reduce cost for low-end systems or to improve energy efficiency
	for battery-powered systems.
\item	ECC error correction might still be done in hardware, avoiding
	the need for precise software-visible exceptions.
\item	Some systems might prefer to immediately shut down in response to
	an uncorrectable ECC error, perhaps due to safety considerations.
	This would entirely avoid the need for software-visible
	ECC-related exceptions.\footnote{
		One motivation for immediate shutdown is that if there
		are uncorrectable errors, there might soon be errors
		that translate one valid bit pattern to another valid
		bit pattern, which could result in a lack of safety.}
\item	Someone might come up with a clever way of correcting ECC errors
	in firmware, avoiding the need for precise software-visible
	exceptions.
\item	As late as early 2024, some GPGPUs have been observed reordering
	earlier loads against later stores.
	Other hardware architects might therefore feel the need to keep
	this option open for their own systems.
\end{enumerate}
